---
title: "Getting and Cleaning Data"
output:
  html_notebook:
      code_folding: none
---

Notes from 'Getting and Cleaning Data', course offered by Johns Hopkins 
University on Coursera at https://www.coursera.org/learn/data-cleaning

***

# Week 3

## Summarising Data

### Getting data from the web

```{r}
if(!file.exists("./Data")){dir.create("./Data")}
fileURL <- "https://data.baltimorecity.gov/api/views/k5ry-ef3g/rows.csv?accessType=DOWNLOAD"
download.file(fileURL, destfile = "./Data/restaurants.csv")
restData <- read.csv("./Data/restaurants.csv", stringsAsFactors = T)
```

### Initial look at the dataset

```{r}
head(restData, n = 3)
tail(restData, n = 3)
summary(restData)
str(restData)
```

### Getting quantiles of quantitative variables

```{r}
quantile(restData$councilDistrict, na.rm = TRUE)
quantile(restData$councilDistrict, probs = c(0.5, 0.75, 0.9))
```

### Making a table from a specific variable

```{r}
table(restData$zipCode, useNA = "ifany")
```

**NB**: 'ifany' adds a cell at the end of the table with the number of missing 
values, if any

```{r}
table(restData$councilDistrict[1:6], restData$zipCode[1:6])
```

### Check for missing values

```{r}
sum(is.na(restData$councilDistrict))
any(is.na(restData$councilDistrict))
all(restData$zipCode > 0) # uh oh we have a negative zip code!
```

### Taking sums

```{r}
colSums(is.na(restData[, 1:6]))
```

### Get values with specific characteristics

```{r}
table(restData$zipCode %in% c("21212"))
restData[restData$zipCode %in% c("21212"), ]
```

### Creating cross tabs

```{r}
data(UCBAdmissions)
DF = as.data.frame(UCBAdmissions)
summary(DF)
```

```{r}
xt <- xtabs(Freq ~ Gender + Admit, data = DF)
xt
```

```{r}
warpbreaks$replicate <- rep(1:9, len = 54)
xt = xtabs(breaks ~., data = warpbreaks)
head(xt)
# to make a FLAT table
ftable(xt)
```

### Getting the size of a dataset

```{r}
fakeData = rnorm(1e5)
object.size(fakeData)
print(object.size(fakeData), units = "Mb")
```


***

## Creating new variables

Commonly you will have to add variables to dataframes such asq:
        - Missingness indicators
        - Cutting up quantitative variables
        - Applying transforms to data
        
### Getting data from the web

```{r}
if(!file.exists("./Data")){dir.create("./Data")}
fileURL <- "https://data.baltimorecity.gov/api/views/k5ry-ef3g/rows.csv?accessType=DOWNLOAD"
download.file(fileURL, destfile = "./Data/restaurants.csv")
restData <- read.csv("./Data/restaurants.csv", stringsAsFactors = T)
```

### Creating sequences

Sequences are often used for creating indexes for data.

```{r}
s1 <- seq(1, 10, by = 2)
s1
s2 <- seq(1, 10, length = 3)
s2
x <- c(1, 3, 8, 25, 100)
seq(along =  x)
```

### Subsetting variables

```{r}
restData$nearMe = restData$neighborhood %in% c("Roland Park", "Homeland")
table(restData$nearMe)
```

### Creating binary variables

```{r}
restData$zipWrong = ifelse(restData$zipCode < 0, TRUE, FALSE)
table(restData$zipWrong, restData$zipCode < 0)
```

### Creating categorical variables from quantitative variables

```{r}
restData$zipGroups = cut(restData$zipCode, breaks = quantile(restData$zipCode))
table(restData$zipGroups)
table(restData$zipGroups, restData$zipCode)
```

```{r}
# without having to specify all the breaks
library(Hmisc)
restData$zipGroups = cut2(restData$zipCode, g = 4)
table(restData$zipGroups)
```
**NB**: Cutting produces *factor variables*.

### Creating factor variables from integer variables

```{r}
restData$zcf <- factor(restData$zipCode)
restData$zcf[1:10]
class(restData$zcf)
```

### Setting the levels of factor variables

```{r}
yesno <- sample(c("yes", "no"), size = 10, replace = TRUE)
yesnofac = factor(yesno, levels = c("yes", "no"))
relevel(yesnofac, ref = "yes")
as.numeric(yesnofac)
```

### Using the `mutate` function in `dplyr`

```{r}
library(Hmisc)
library(dplyr)
restData2 = mutate(restData, zipGroups = cut2(zipCode, g = 4))
table(restData2$zipGroups)
```

### Common transforms of data

- `abs(x)` get the absolute value
- `sqrt(x)` square root
- `ceiling(x)` round up to nearest integer
- `floor(x)` round down to nearest integer
- `round(x, digits = n)` round to *n* digits
- `signif(x, digits = n)` rount to *n* significant figures
- `cos(x)`, `sin(x)`, etc.
- `log(x)`, `log2(x)`, `log10(x)`, etc.
- `exp(x)` exponentiate

***

## Reshaping Data

Often data you load into R will not necessarily by *tidy*, and will need 
*reshaping*  to be made tidy. Recall that with tidy data:

1. Each variable forms a column,
2. Each observation forms a row,
3. Each table stores data about one kind of observation.

```{r}
head(mtcars)
```

As we can see, the above table does *not* confirm to tidy data principles

### Melting the dataset

*Melting* a dataset means that we reshape the data so that each row is a *unique id-variable combination*.

```{r}
library(reshape2)
mtcars$carname <- rownames(mtcars)
carMelt <- melt(mtcars, id = c("carname", "gear", "cyl"),
                measure.vars = c("mpg", "hp"))
head(carMelt, n = 3)
tail(carMelt, n = 3)
```

We have reshaped the dataset so that:

- *carnames* is now a variable
- *variables* is a single column for both mpg and hp

### Recasting the data set

We can now recast the melted data into any shape that we want, *e.g.*:

```{r}
cylData <- dcast(carMelt, cyl ~ variable, mean)
cylData
```

### Averaging values

```{r}
head(InsectSprays)
# within each value of 'spray', sum up the 'count'
tapply(InsectSprays$count, InsectSprays$spray, sum)
```

Alternatively we can *split, apply, combine*:

```{r}
spIns = split(InsectSprays$count, InsectSprays$spray)
spIns
sprCount = lapply(spIns, sum)
sprCount
unlist(sprCount)
```

Or we can use `plyr`:

```{r}
library(plyr)
ddply(InsectSprays, .(spray), summarise, sum = sum(count))
```

***

## Managing data frames with `dplyr`

`dplyr` is a package specifically designed for manipulating data frames.

### Verbs:

- `select`: return a subset of the columns of a data frame
- `filter`: extract a subset of rows from a data frame based on logical conditions
- `arrange`: reorder rows of a data frame
- `rename`: rename variables in a data frame
- `mutate`: add new variables, columns or transform existing variables
- `summarise`: generate summary statistics of different variables 

### Properties:

- The first argument is always a data frame.
- The following arguments  describe what to do with the data frame.
   - **NB**: You can refer to columns directly without using the `$` operator.
- The result is always a *new* data frame.

**NB**: Data frames must be *properly formatted* and *annotated* for these functions to be useful.

### Examples of basic verbs:

#### `select`
```{r}
library(dplyr)
chicago <- readRDS("./Data/chicago.rds")
names(chicago)
head(select(chicago, city:dptp)) # select just by column names
head(select(chicago, -city:dptp)) # take negative of previous selection
```

#### `filter`
```{r}
chic.f <- filter(chicago, pm25tmean2 > 30) # filter by logical test on one column
head(chic.f)
chic.f <- filter(chicago, pm25tmean2 > 30 & tmpd > 80) # filter by logical test on multiple columns
head(chic.f)
```

#### `arrange`
```{r}
chicago <- arrange(chicago, date) # reorder rows by date variable
head(chicago)
chicago <- arrange(chicago, desc(date)) # reorder rows in descebding order by date
head(chicago)
```

#### `rename`
```{r}
chicago <- rename(chicago, pm25 = pm25tmean2, dewpoint = dptp) # rename columns
head(chicago)
```

#### `mutate`
```{r}
# add a new column showing variation from the mean of pm25
chicago <- mutate(chicago, pm25detrend = pm25 - mean(pm25, na.rm = TRUE)) 
head(select(chicago, pm25, pm25detrend))
```

#### `group_by`
```{r}
# create a factor variable (tempcat) which labels days as either hot or cold
chicago <- mutate(chicago, tempcat = factor(1 * (tmpd > 80), 
                                            labels = c("cold", "hot")))
hotcold <- group_by(chicago, tempcat) # create a new table based on tempcat
summarise(hotcold, pm25 = mean(pm25, na.rm = T), 
          o3 = max(o3tmean2), 
          no2 = median(no2tmean2))
# create a new column with the year extracted from the date
chicago <- mutate(chicago, year = as.POSIXlt(date)$year + 1900) 
years <- group_by(chicago, year) # group by year and summarise
summarise(years, pm25 = mean(pm25, na.rm = T), 
          o3 = max(o3tmean2), 
          no2 = median(no2tmean2))
```

### Using the `%>%` operator to join verbs
```{r}
chicago %>%
        mutate(month = as.POSIXlt(date)$mon + 1) %>%
        group_by(month) %>%
        summarise(pm25 = mean(pm25, na.rm = T), 
                  o3 = max(o3tmean2), 
                  no2 = median(no2tmean2))
```

***

## Merging Data

```{r}
if(!file.exists("./Data")){dir.create("./Data")}
fileUrl1 = "https://raw.githubusercontent.com/DataScienceSpecialization/courses/master/03_GettingData/04_01_editingTextVariables/data/reviews.csv"
fileUrl2 = "https://raw.githubusercontent.com/DataScienceSpecialization/courses/master/03_GettingData/04_01_editingTextVariables/data/solutions.csv"
download.file(fileUrl1, "./Data/reviews.csv", mode = "wb")
download.file(fileUrl2, "./Data/solutions.csv", mode = "wb")
```

```{r}
reviews = read.csv("./Data/reviews.csv")
solutions = read.csv("./Data/solutions.csv")
head(reviews, 2)
head(solutions, 2)
```
**NB**: reviews dataset has a variable called "solution_id" which matches to 
variable "id" in the solution dataset

### `merge()`

Important parameters: 

- `x, y`: data frames
- `by, by.x, by.y`: what to merge by
- `all`: when TRUE, if there is a value unique to one dataset, add row with NAs

**NB**: By default, we will merge by *all* columns that have a common name 
between the two datasets. This may not be what you want!

```{r}
mergedData = merge(reviews, solutions, 
                   by.x = "solution_id", by.y = "id", 
                   all = TRUE) # merge by combining x$solution_id with y$id
head(mergedData)
```

### `plyr::join`

Faster, but requires that you have a *common variable* to join by. 
However if you have multiple data frames it is much easier than using `merge()`.

```{r}
library(plyr)
df1 = data.frame(id = sample(1:10), x = rnorm(10))
df2 = data.frame(id = sample(1:10), y = rnorm(10))
df3 = data.frame(id = sample(1:10), z = rnorm(10))
dfList = list(df1, df2, df3)
join_all(dfList)
```

***


