---
title: "R Programming"
output:
  html_notebook:
      code_folding: none
---

This notebook contains notes taken from the Coursera course [R Programming](https://www.coursera.org/learn/r-programming) created by Johns 
Hopkins University.

***

## Week 1

### Reading and writing data into R

- `read.table, write.table` for reading/writing tabular data
- `readLines, writeLines` for reading/writing lines of a text file
- `source, dump` or `dget, dput` for reading/writing R code files

#### Tabular formats

`read.table` arguments:

- `file`, for filename/connection
- `header`, logical, indicates if there is a header line
- `sep`, string, indicates the column separator
- `colClasses`, vector, indicates class of each column (if blank, R will attempt
to calculate)
- `nrows`, number of rows (if blank, R will calculate)
- `comment.char`, string, indicates comment character (default: `,` for 
`read.csv`)
- `skip`, number of lines to skip from the beginning (default: `#`)
- `stringsAsFactors` logical, should character variables be encoded as factors?

When reading large datasets:

- read [help page](https://www.rdocumentation.org/packages/utils/versions/3.6.2/topics/read.table) for `read.table`
- know how much memory is available, what other applications or users are using 
resources, what the OS is (32-bit or 64-bit?)
- roughly calculate amount of memory required to store the dataset:
- $\text{nrows} \times \text{ncolumns} \times 8 \text{ bytes/numeric}$
- set `comment.char = ""` if there are no comments
- set `nrows` so R can assign appropriate amount of memory (use `wc` in bash)
- specify `colClasses` so R doesn't have to work each column class out. To work 
out classes of each column you can just check the first ~100 rows and loop over 
each column using `sapply`:

```{r}
initial <- read.table("datatable.txt", nrows = 100)
classes <- sapply(initial, class)
tabAll <- read.table("datatable.txt", colClasses = classes)
```

#### Textual formats

- `dump` or `dput` can preserve the *metadata*  associated with specific columns
or rows
- `dput` will *deparse* R objects, can pull metadata for us
- `dump` can be used on multiple R objects
- tend to work better with version control programs (*e.g.* git)
- can be safer for long-term storage as harder to repair corruption
- not very space-efficient

#### Interfacing with the outside world

- `file`, opens a connection to a file
- `gzfile`, opens up to a file compressed with gzip
- `url`, opens a connection to a webpage

Uses of connections:

- reading or writing specific lines of a text file or website:

```{r}
con1 <-gzfile("words.gz")
x <- readLines(con1, 10)

con2 <-url("http://www.sample.com")
y <- readLines(con2, 10)
```

### Subsetting data

#### Operators:

- `[]` always returns an object of the *same class* as the original, can be used
to select *multiple elements*
- `[[]]` extracts elements of a list or data frame, can only extract a *single* 
*element*, class of returned object will not necessarily be the same
- `$` extracts *named* elements of a list or data frame

#### Subsetting vectors:

```{r}
x <- c("a", "b", "c", "c", "d", "a")
x[1:4] # values in [] are a numeric index
x[x > "a"] # value in [] is a logical index
```

#### Subsetting lists:

```{r}
x <-list(foo = 1:4, bar = 0.6)
x[1] # will generate a list
x[[1]] # will generate a sequence (not a list)
x$bar # will get the object "bar"
```

**NB:** `[[]]` operator can also be used with *computed indices* which are put 
into variables, whereas `$` can only be used with *literal* names.

#### Subsetting matrices:

```{r}
x <- matrix(1:6, 2, 3) # creates a matrix
x 
x[1, 2] # subsets object in first row, second column
x[, 2] # subsets entire second column
```

**NB:** By default, when a single element, column or row is retrieved with `[]`,
it will return a *vector* of length 1. To return a 1×1 *matrix*, set 
`drop = FALSE`.

#### Subsetting with names:

Partial matching of names is possible using the `[[]]` and `$` operators.

```{r}
x <- list(aardvark = 1:5)
x$a
x[["a"]] # expects an exact match
x[["a", exact = FALSE]] # will extract a partial match
```

#### Removing missing data:

```{r}
x <- c(1, NA, NA, 4, NA, 5)
bad <- is.na(x) # vector of NA values in x
x[!bad] # returns vector of values in x that are *not* NA

y <- c("a", "b", NA, "d", NA, "f")
good <- complete.cases(x,y) # vector where *both* elements are *not* NA
y[good]

airquality[1:6, ] # this table has missing values
good <- complete.cases(airquality) # vector of rows with no missing data
airquality[good, ][1:6, ] # now only includes rows without missing values
```

### Vectorised operations 

R's ability to perform vectorised operations reduces your need to write `for` or
`while` loops for simple calculations.

```{r}
x <- 1:4
y <- 6:9
x + y # adds nth value from x to equivalent value of y in parallel
x > 2 # compares all values in x to 2

x <- matrix(1:4, 2, 2)
y <- matrix(rep(10,4), 2, 2)
x * y # this will perform an *element-wise multiplication*
x %*% y # this will perform a *true matrix multiplication* 
```

***

## Week 2

### Control Structures

- `if`, `else`: test a condition
- `for`: execute a loop a fixed number of times
- `while`: execute a loop while a condition is true (*can be infinite if* 
*condition is never true*)
- `repeat`: execute an infinite loop
- `break`: break out of a loop entirely
- `next`: skip an iteration in a loop
- `return`: exit a function and return a value

For example:
```{r}
if (x > 3) {
  y <- 10
} else {
  y <- 0
}

# or we can assign the entire construct to y

y <- if(x > 3) {
  10
} else {
  0
}
```

### Writing functions

- created using `function()`
- treated as objects, therefore can be passed as arguments to other functions 
and can be *nested* inside one another
- return value of a function is the last expression to be evaluated

#### Arguments:

- *named arguments* can have *default values*
- *formal arguments* are included in the function's definition
- `formals` function will return a list of all formal arguments in a function
- not every function uses all formal arguments, some can be missing or will 
have default values
- arguments can be matched *positionally* (by position in the function 
definition) or *by name* (with `=`)
- **NB**: when an argument is matched *by name* it is "taken out" of the 
argument list, the remaining arguments will be matched *positionally*
- **NB**: function arguments can also be *partially* matched with the following 
logic:
1. look for exact match for named argument
2. look for partial match
3. perform positional match
- the ellipsis (...) argument can be used to indicate a *variable number of 
arguments* that are passed on to other functions
- **NB**: any functions after the ellipsis argument must be explicit and 
*cannot* be partically matched

#### Lazy evaluation:
R will only evaluate function arguments *as needed*, for example:

```{r}
f <- function(a, b) {
  print(a)
  print(b)
}
f(45)
```
Note how the error only occured *after* the first argument was evaluated and the
first value was printed out.

#### Examples of writing functions:
```{r}
add2 <- function(x, y){
  x + y
}

# once we've defined the function, we can call it like so:

add2(5, 6) 
```

```{r}
above_n <- function(x, n = 10) {
  use <- x > n # work out which elements are > n (default to 10)
  x[use] # return those values
}

above_n(1:20, 12)
above_n(1:20) # without n input, will default to n = 10
```

```{r}
columnmean <- function(y, removeNA = TRUE) {
  nc <- ncol(y) # count the columns in df/matrix
  means <- numeric(nc) # initialise empty vector of same length to store means
  for(i in 1:nc) { # for each number from 1 to [number of columns]
    means[i] <- mean(y[, i], na.rm = removeNA) # add mean of the ith column 
    # remove missing values if removeNA = TRUE.
  }
  means # return the vector of means
}

columnmean(airquality) # test on a dataset
```

**NB**: Generally it is best to save functions as R scripts and source them into
R. Here they are included in the notebook for simplicity's sake.

### Coding Standards for R
1. Always use plain text
2. Indent your code (*e.g.* 4 to 8 char)
3. Limit code width (*e.g.* to 80 columns)

This will make you limit sizes of functions, and will make your code more
readable to yourself and to others and easier to debug if things go wrong.

### Dates and Times in R
Dates and times are represented in special classes:

- `Date` for dates, stored as number of *days* since 1970-01-01
- `POSIXct` and `POSIXlt` for times, stored as number of *seconds* since 
1970-01-01
  - `POSIXct` stores information as a large integer, useful for storing times
in data frames, etc.
  - `POSIXlt` stores information as a *list*, can include other useful info.
- a number of generic functions give us other info when applied to dates/times:
  - `weekdays` gives day of the week
  - `months` gives the month
  - `quarters` gives the annual quarter (Q1 - Q4)
- the `strptime` function can convert dates written in strings in different
string formats to `Date`/`POSIXlt`format
- you can perform mathematical operations on dates/times *so long as they are*
*in the same format*

Example of converting to date:
```{r}
x <- as.Date("1992-04-11")
x # looks like a character string
class(x) # actually a Date class
unclass(x) # raw data stored as a number of days
```

Example of mathematical operations with dates:
```{r}
x <- as.Date("2012-03-01")
y <- as.Date("2012-02-28")

x - y # Find difference (in days) between two dates.
```
**NB**: R will take care of complicating facors such as leap years, time zones, 
etc. automatically.

***

## Week 3 

### Loop functions

- `lapply`: Loop over a list, apply a function to each element
- `sapply`: Same as `lapply` but try to simplify the result
- `apply`: Apply a function over the margins of an array (*e.g.* summarise
matricies, etc.)
- `tapply`: Apply a function over subsets of a vector
- `mapply`: Multivariate version of `lapply`
- `split`: Splits objects into sub-spieces, often used in conjunction with the
above functions

#### lapply

`lapply` takes three arguments:

  1. a list (`x`)
  2. a function (`FUN`)
  3. other arguments (`...`)
  
**NB:** If `x` is not a list, it will be coerced using `as.list`. 
Output will *always be a list*.

```{r}
lapply
```

For example:

```{r}
x <- list(a = 1:5, b = rnorm(10))
lapply(x, mean)
```

```{r}
x <- 1:4
lapply(x, runif, min = 0, max = 10)  # call runif on each element of x

```

```{r}
x <- list(a = matrix(1:4, 2, 2), b = matrix(1:6, 3, 2))
# suppose we want to extract the first column from each matrix?
# we can write an *anonymous function* which will only exist within this loop:
lapply(x, function(element) element[, 1])
```

### sapply

`sapply` tries to *simplify* `lapply` output, *e.g.*:

- If the result is a list where every element has length = 1, return a *vector*.
- If the result is a list where every element has equal length > 1, return a 
*matrix*.
- If the result can't be simplified, just return a list.

### apply

`apply` is usually used to evaluate a (often anonymous) function over the
margins of an array. 

- Usually this means applying a function to the *rows or columns of a matrix*.
- Can also be used with general arrays, *e.g.* mean of an array of matricies.
- Not really faster than writing a loop, but works in one line, very useful
on the command line.

```{r}
str(apply)
```

- `X` is an *array*.
- `MARGIN` is an integer vector indicating which *margins* should be "retained".
- `FUN` is the function to be applied.
- `...` is for other arguments passed to `FUN`.

For example:

```{r}
x <- matrix(rnorm(200), 20, 10) # a matrix of 20 x 10
apply(x, 2, mean) # calculate the mean of each column of the matrix
apply(x, 1, sum) # calculate the sum of each row of the matrix
```

**NB**: The *margin* here indicates the *dimension*  of the matrix we want 
to 'keep'. For sums and means of matricies there are actually some shortcuts:

- `rowSums = apply(x, 1, sum)`
- `rowMeans = apply(x, 1, mean)`
- `colSums = apply(x, 2, sum)`
- `colMeans = apply(x, 2, mean)`

Other examples:
```{r}
x <- matrix(rnorm(200), 20, 10)
apply(x, 1, quantile, probs = c(0.25, 0.75)) # get quantiles of each row
```

```{r}
a <- array(rnorm(2 * 2 * 10), c(2, 2, 10)) # 3-dimensional array
apply(a, c(1, 2), mean) # get the average of these 2x2 matricies, keep dim 1 & 2
rowMeans(a, dims = 2) # can also use rowMeans
```

### mapply

`mapply` applies a function *in parallel* over a set of arguments, for example 
if we have *two* lists we want to apply a function over.

```{r}
str(mapply)
```

- `FUN` is the function to apply.
- `...` contains the list of arguments to apply over.
- `MoreArgs` is a list of other arguments of `FUN`
- `SIMPLIFY`indicates whether to simplify result like in `sapply`.

**NB**: The function you use must have at least the same number of arguments
as items in the list of arguments you pass to it.

For example:
Instead of typing `list(rep(1, 4), rep(2, 3), rep (3, 2), rep (4, 1)) we can do:

```{r}
mapply(rep, 1:4, 4:1)
```

### tapply

`tapply` applies functions over *subsets of a vector*. 

```{r}
str(tapply)
```

- `X` is a *vector*.
- `INDEX` is a factor/list of factors, indicating your *subset*.
- `FUN` is the function to be applied.
- `...` contains other arguments passed to `FUN`.
- `simplify` is same as in `sapply`.

For example:
```{r}
x <- c(rnorm(10), runif(10), rnorm(10, 1)) # our vector of three distributions
f <- gl(3, 10) # our factor, with three levels, each repeated 10 times
tapply(x, f, mean) # get the mean of each of the three groups in x
```

### split

`split` takes a vector or other object and splits it into groups determined by 
a factor or list of factors. It will always return a *list*.

It is similar to `tapply` but without applying the summary statistics.

```{r}
str(split)
```

- `x` is an object.
- `f` is a factor/list of factors.
- `drop` indicates whether *empty* factor levels should be dropped.

For example:

```{r}
x <- c(rnorm(10), runif(10), rnorm(10, 1))
f <- gl(3, 10)
split(x, f)
```

Here we see `split` has split out vector of three distributions into a 
list with 3 parts. We can also split a *data frame*:

```{r}
library(datasets)
s <- split(airquality, airquality$Month) # split data frame by Month (-> factor)
sapply(s, function(x) colMeans(x[, c("Ozone", "Solar.R", "Wind")], na.rm=TRUE))
```

We can also split on *multiple levels*:

```{r}
x <- rnorm(10) # vector
f1 <- gl(2, 5) # first factor, 2 levels, each repeated 5 times
f2 <- gl(5, 2) # second factor, 5 levels, each repeated 2 times
str(split(x, list(f1, f2), drop = TRUE)) # automatically gets the interactions
```

### Debugging

When there is a problem, R can return certain notifications:

- `message`: generic notification, execution of function continues.
- `warning`: indication something is wrong, execution of function continues.
- `error`: indication of fatal problem, execution is stopped.
  - these are all a `condition`: indicates that something unexpected occurred.
  
Questions to ask yourself when debugging:

- What was the actual input?
- How was the function called?
- What were you expecting? Output, messages, other result?
- What did you get instead?
- How does what you got differ from what you expected?
- Were your expectations correct in the first place?
- Can you reproduce the problem exactly?

Base R has certain tools built in for debugging:

- `traceback`: prints out the *function call stack* after an error occurs.
- `debug`: flags a function for "debug" mode, allowing you to step through
execution *one step at a time*.
- `browser`: suspends the execution of a function *wherever it is called*
and puts it in "debug" mode.
- `trace`: lets you insert debugging code into a function without actually 
editing the function itself.
- `recover`: allows you to modify the error behaviour, so you can browse the 
function call stack, (*e.g.* not go straight back to console).

***





