---
title: "R Programming"
output: html_notebook
---

This notebook contains notes taken from the Coursera course [R Programming](https://www.coursera.org/learn/r-programming) created by Johns Hopkins University.

## Week 1

### Reading and writing data into R

- `read.table, write.table` for reading/writing tabular data
- `readLines, writeLines` for reading/writing lines of a text file
- `source, dump` or `dget, dput` for reading/writing R code files

#### Tabular formats

`read.table` arguments:

- `file`, for filename/connection
- `header`, logical, indicates if there is a header line
- `sep`, string, indicates the column separator
- `colClasses`, vector, indicates class of each column (if blank, R will attempt to calculate)
- `nrows`, number of rows (if blank, R will calculate)
- `comment.char`, string, indicates comment character (default: `,` for `read.csv`)
- `skip`, number of lines to skip from the beginning (default: `#`)
- `stringsAsFactors` logical, should character variables be encoded as factors?

When reading large datasets:

- read [help page](https://www.rdocumentation.org/packages/utils/versions/3.6.2/topics/read.table) for `read.table`
- know how much memory is available, what other applications or users are using resources, what the OS is (32-bit or 64-bit?)
- roughly calculate amount of memory required to store the dataset:
     - $\text{nrows} \times \text{ncolumns} \times 8 \text{ bytes/numeric}$
- set `comment.char = ""` if there are no comments
- set `nrows` so R can assign appropriate amount of memory (use `wc` in bash)
- specify `colClasses` so R doesn't have to work each column class out. To work out classes of each column you can just check the first ~100 rows and loop over each column using `sapply`:

```{r}
initial <- read.table("datatable.txt", nrows = 100)
classes <- sapply(initial, class)
tabAll <- read.table("datatable.txt", colClasses = classes)
```

#### Textual formats

- `dump` or `dput` can preserve the *metadata*  associated with specific columns or rows
  - `dput` will *deparse* R objects, can pull metadata for us
  - `dump` can be used on multiple R objects
- tend to work better with version control programs (*e.g.* git)
- can be safer for long-term storage as harder to repair corruption
- not very space-efficient

#### Interfacing with the outside world

- `file`, opens a connection to a file
- `gzfile`, opens up to a file compressed with gzip
- `url`, opens a connection to a webpage

Uses of connections:

- reading or writing specific lines of a text file or website:

```{r}
con1 <-gzfile("words.gz")
x <- readLines(con1, 10)

con2 <-url("http://www.sample.com")
y <- readLines(con2, 10)
```

### Subsetting data

#### Operators:

- `[]` always returns an object of the *same class* as the original, can be used to select *multiple elements*
- `[[]]` extracts elements of a list or data frame, can only extract a *single element*, class of returned object will not necessarily be the same
- `$` extracts *named* elements of a list or data frame

#### Subsetting vectors:

```{r}
x <- c("a", "b", "c", "c", "d", "a")
x[1:4] # values in [] are a numeric index
x[x > "a"] # value in [] is a logical index
```

#### Subsetting lists:

```{r}
x <-list(foo = 1:4, bar = 0.6)
x[1] # will generate a list
x[[1]] # will generate a sequence (not a list)
x$bar # will get the object "bar"
```

**NB:** `[[]]` operator can also be used with *computed indices* which are put into variables, whereas `$` can only be used with *literal* names.

#### Subsetting matrices:

```{r}
x <- matrix(1:6, 2, 3) # creates a matrix
x 
x[1, 2] # subsets object in first row, second column
x[, 2] # subsets entire second column
```

**NB:** By default, when a single element, column or row is retrieved with `[]`, it will return a *vector* of length 1. To return a 1×1 *matrix*, set `drop = FALSE`.

#### Subsetting with names:

Partial matching of names is possible using the `[[]]` and `$` operators.

```{r}
x <- list(aardvark = 1:5)
x$a
x[["a"]] # expects an exact match
x[["a", exact = FALSE]] # will extract a partial match
```

#### Removing missing data:

```{r}
x <- c(1, NA, NA, 4, NA, 5)
bad <- is.na(x) # vector of NA values in x
x[!bad] # returns vector of values in x that are *not* NA

y <- c("a", "b", NA, "d", NA, "f")
good <- complete.cases(x,y) # vector of positions where *both* elements are *not* NA
y[good]

airquality[1:6, ] # this table has missing values
good <- complete.cases(airquality) # vector of rows with no missing data
airquality[good, ][1:6, ] # now only includes rows without missing values
```

### Vectorised operations 

R's ability to perform vectorised operations reduces your need to write `for` or `while` loops for simple calculations.

```{r}
x <- 1:4
y <- 6:9
x + y # adds nth value from x to equivalent value of y in parallel
x > 2 # compares all values in x to 2

x <- matrix(1:4, 2, 2)
y <- matrix(rep(10,4), 2, 2)
x * y # this will perform an *element-wise multiplication*
x %*% y # this will perform a *true matrix multiplication*
```
